Methods

### 2022-05-15 article/abs.Rmd
revise intro,bibs

# ATBC *(400 wds)*

## REVIEW CRITERIA
​
All abstracts will be evaluated and scored by the ATBC2022 Scientific Committee. The Scientific Committee may accept, decline, or offer an opportunity for a revised abstract submission. Proposals will be assessed using the following criteria:

- *Scientific merit*: Likelihood of promoting a significant advancement in our understanding of tropical biology and conservation, innovative or interdisciplinary approaches, or likelihood of novel conceptual synthesis.

- *Broader impacts*: Integration of science with conservation issues, impact on policy, and potential utility for conservation, among others.

- *Appeal to participants of ATBC2022*: Abstract should target topics that will be of interest to the ATBC community.

- Relevance to the meeting's theme: We encourage submission of abstracts that are relevant to the meeting’s theme.

- A clearly written and compelling abstract.

# Draft = Habitat edges affect tree diversity more than biomass regeneration in a wet tropical forest

## Introduction / Background / Justification *(50-80 wds)*

About half of all forests are tropical and secondary, making tropical forest regeneration processes integral to the future of forests globally.

However, syntheses find forest regeneration to be quick yet inconsistent--stand biomass and taxonomic richness recovers almost completely in about half a century, but relative abundances may take centuries to recover or instead may diverge entirely.

Given that three quarters of all forests are within a km of a habitat edge, nearby habitat edges also likely affect biomass and taxonomic regeneration trends, as well as management history legacies.

In wet tropical forests specifically, succession tends to depend on light availability, compared to water availability and hydraulic traits like more open or dry forests, which may also extend to mediate edge effects.

## Objective(s)/Hypothesis(es) *(50-80 wds)*

This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge in a former timber plantation context, and
addresses the hypothesis that old plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and that habitat edges affect regenerating community composition.

## Methods *(50-80 wds)*

A 20 ha timber plantation abandoned ca 20 years ago surrounded by primary forest on two sides and a main service road on a third was censused for over a thousand trees using random stratified sampling plots representing 5% of total parcel area along a 300 m edge distance gradient in the Osa Peninsula, SW Costa Rica, and analysis used relevant databases from the literature.

## Results *(50-80 wds)*

As distance from primary forest edge increased, biomass and stem and wood density tended to increase, with ca 10% variation explained, while canopy light and stand height tended to stay the same.

Stand tree richness also tended to increase, but diversity decreased steeply and non-linearly, and composition varied notably.

Tree taxa associated with both early and late successional stages increased, but biomass by dispersal mode did not tend to change.

## Implications/Conclusions *(50-80 wds)*

Overall this study supports main ideas about forest regeneration that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future with timber and diversity trade-offs based on local and regional management activity.

## Keywords (10)

forest
secondary
regeneration
edge
biomass
diversity
community
composition
function
trait

# SACNAS

About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests. 
Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely. 
Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, medaited by shade tolerance and dispersal traits, along with management history. 
This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge, and hypothesizes that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition. 
An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature. 
As distance from primary forest edge increased, biomass and stem and wood density tended to increase, with ca 10% variation explained, while canopy light and stand height tended to stay the same. 
Stand tree richness also tended to increase, but diversity decreased steeply and non-linearly, and composition varied notably. 
Finally, tree taxa associated with both early and late successional stages increased, but biomass by dispersal mode did not tend to change. 
Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.

>>>>

<!-- # ATBC *(400 wds)* -->

<!-- ## REVIEW CRITERIA -->
<!-- ​ -->
<!-- All abstracts will be evaluated and scored by the ATBC2022 Scientific Committee. The Scientific Committee may accept, decline, or offer an opportunity for a revised abstract submission. Proposals will be assessed using the following criteria:  -->

<!-- - *Scientific merit*: Likelihood of promoting a significant advancement in our understanding of tropical biology and conservation, innovative or interdisciplinary approaches, or likelihood of novel conceptual synthesis.  -->

<!-- - *Broader impacts*: Integration of science with conservation issues, impact on policy, and potential utility for conservation, among others.  -->

<!-- - *Appeal to participants of ATBC2022*: Abstract should target topics that will be of interest to the ATBC community.  -->

<!-- - Relevance to the meeting's theme: We encourage submission of abstracts that are relevant to the meeting’s theme. -->

<!-- - A clearly written and compelling abstract. -->

<!-- # Draft = Habitat edges affect tree diversity more than biomass regeneration in a wet tropical forest -->

<!-- ## Introduction / Background / Justification *(50-80 wds)* -->

<!-- About half of all forests are tropical and secondary, making tropical forest regeneration processes integral to the future of forests globally. -->

<!-- However, syntheses find forest regeneration to be quick yet inconsistent--stand biomass and taxonomic richness recovers almost completely in about half a century, but relative abundances may take centuries to recover or instead may diverge entirely. -->

<!-- Given that three quarters of all forests are within a km of a habitat edge, nearby habitat edges also likely affect biomass and taxonomic regeneration trends, as well as management history legacies. -->

<!-- In wet tropical forests specifically, succession tends to depend on light availability, compared to water availability and hydraulic traits like more open or dry forests, which may also extend to mediate edge effects. -->

<!-- ## Objective(s)/Hypothesis(es) *(50-80 wds)* -->

<!-- This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge in a former timber plantation context, and -->
<!-- addresses the hypothesis that old plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and that habitat edges affect regenerating community composition. -->

<!-- ## Methods *(50-80 wds)* -->

<!-- A 20 ha timber plantation abandoned ca 20 years ago surrounded by primary forest on two sides and a main service road on a third was censused for over a thousand trees using random stratified sampling plots representing 5% of total parcel area along a 300 m edge distance gradient in the Osa Peninsula, SW Costa Rica, and analysis used relevant databases from the literature. -->

<!-- ## Results *(50-80 wds)* -->

<!-- As distance from primary forest edge increased, biomass and stem and wood density tended to increase, with ca 10% variation explained, while canopy light and stand height tended to stay the same. -->

<!-- Stand tree richness also tended to increase, but diversity decreased steeply and non-linearly, and composition varied notably. -->

<!-- Tree taxa associated with both early and late successional stages increased, but biomass by dispersal mode did not tend to change. -->

<!-- ## Implications/Conclusions *(50-80 wds)* -->

<!-- Overall this study supports main ideas about forest regeneration that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future with timber and diversity trade-offs based on local and regional management activity. -->

<!-- ## Keywords (10) -->

<!-- forest -->
<!-- secondary -->
<!-- regeneration -->
<!-- edge -->
<!-- biomass -->
<!-- diversity -->
<!-- community -->
<!-- composition -->
<!-- function -->
<!-- trait -->

<!-- # SACNAS -- Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest -->

<!-- About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests.  -->
<!-- Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely.  -->
<!-- Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, mediated by shade tolerance and dispersal traits, along with management history. -->
<!-- This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge, and hypothesizes that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition. -->
<!-- An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature. -->
<!-- As distance from primary forest edge increased, biomass and stem and wood density tended to increase, with ca 10% variation explained, while canopy light and stand height tended to stay the same. -->
<!-- Stand tree richness also tended to increase, but diversity decreased steeply and non-linearly, and composition varied notably.  -->
<!-- Finally, tree taxa associated with both early and late successional stages increased, but biomass by dispersal mode did not tend to change. -->
<!-- Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity. -->

# ESA -- Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest

<!-- # Background/Question/Methods -->

About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests. 
Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely. 
Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, mediated by shade tolerance and dispersal traits, along with management history.
This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge, and hypothesizes that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature.

<!-- # Results/Conclusions -->

As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ca 10% variation explained, while biomass and canopy light tended to stay the same.
Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, and taxonomic composition varied notably. 
Finally, tree taxa associated with both early and late successional stages decreased significantly, but biomass by dispersal mode did not tend to change.
Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.

### 2022-05-15 article/disc.Rmd
revise intro,bibs

---
title: "Discussion"
output: html_notebook
---

@@

>>>>

---
title: "Discussion"
output: html_notebook
bibliography: ../../library.bib
link-citations: TRUE
---

@@

The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density *[@césar21]*, and could therefore require a switch in focal timber species *[@karsten14]*.

### 2022-05-15 article/intro.Rmd
revise intro,bibs

---
title: "Introduction"
output: html_notebook
---

1.  importance of reforestation/regeneration/regrowth

2.  patterns of reforestation

    1.  general

    2.  tropical

3.  things that affect reforestation

    1.  edge effects

    2.  plantation structure?

4.  [for NGO--start biomass census]

    1.  Does leaving the plantation structure help reforestation?

        1.  shade

        2.  edge effects (i.e. buffer)

    2.  How does the forest regenerate by distance to primary forest?

        1.  how much [effort] do we need to re-seed? / how much can we leave to nature?

Global forest and landscape regeneration is a key conservation and climate change mitigation strategy, adopted internationally (\@UN14 Paris, Bonn Challenge).

-   Tropical forests store most terrestrial biomass ( @pan2013 ) and biodiversity (CITE, but are losing functions, especially in the neotropics ( @hubau2020 ), as most tropical forests become secondary (\@FAO20.

Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.

-   Neotropical secondary forests accumulate biomass \>10x faster than old-growth forests ( @poorter2016 ), and offset \~10% of carbon emissions from Amazonian deforestation ( @smith2020 ).
-   They also recover taxonomic richness quite quickly, averaging 80% of old-growth forest levels in 20y ( @rozendaal2019 ).
-   However, secondary neotropical forests diverge widely in taxonomic composition ( @norden2015 ).

Divergent communities among secondary forests may ultimately warrant more localized conservation efforts, studying relative effects of generalized community processes.

Half of all forests lie half of a kilometer from an ecotone ( @haddad2015 ), suggesting that edge effects from adjacent habitats are widespread aspects of secondary forest regeneration.

-   Tropical forests lose 22% carbon along edges \>100m, in part due to shorter trees and less leaf mass ( @ordway2020 ).

Here we report the edge effects of an abandoned wet neotropical timber plantation.

-   We hypothesized that remnant timber trees would mediate forest regrowth by maintaining uniform shade that would suppress shade-intolerant / pioneer trees / taxa and allow for more diversity.

    -   supported by inference?!

Forest regeneration has the potential to store \#\#C and preserve \#\# diversity globally in the coming decades (UN, Bonn challenge).

At local scales, regeneration is strongly affected by landscape [ecological] heterogeneity as well as management history.

However, patterns of forest regeneration remain unclear, especially at more local scales (\< \~50 ha).

Syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (\~50-80%, compared to old-growth forests) rather quickly (decades, \< 20 years), mainly with respect to tree height, above-ground biomass (Poorter et al 2016), and species richness, especially in wetter regions (Rozendaal et al 2019, Onyekwelu & Olabiwonnu 2016).

However, while restoring species richness may be enough to maintain some ecosystem functions like carbon storage (Liu et al 2018), it does not guarantee the restoration of species composition.

Instead, changes to species composition likely last up to centuries, showing a strong legacy of priority effects from pioneering, early-successional communities that are mostly fast-growing, low-wood-density, and shade-intolerant (Tabarelli et al 2014, Tabarelli & Lopes 2008).

forests show legacy effects of common human activity, like logging and intensive agriculture/agroforestry/silviculture (FAO 2020).

[High] Regional biodiversity [can] affects/increases the uncertainty about the local occurrences and outcomes of specific species interactions,

such as after dispersal and/or niche overlap [CITE-niche, neutral theory].

[At the same time,] Diverse local management activity modifies demographic states/profiles of various taxa's population dynamics.

For example, active timber plantations strongly promote the population growth of focal taxa at the expense of non-focal taxa.

After abandonment, however, the ecosystem likely reflects how continuous regional ecological processes look different given past management.

such as seed dispersal from nearby primary forest edges.

[These] Interactions between background ecology and human management ultimately have both local and general implications for the potential conservation/stability of ecosystem services/functions

like biomass/carbon storage and biodiversity maintenance,

as well as manager livelihood.

Slower changes in composition can also be influenced by seed banks, herbivore communities, and dispersal from the surrounding landscape, all of which may affect diverging restoration trajectories among tropical forests (Nordon et al 2015, Christensen 2014, Munroe et al 2013, Chazdon et al 2007).

but lagging changes in composition may contribute to reduced effects of diversity on other forest functions (Griess et al 2012).

Restoring species composition more quickly is likely a place where active forest managers can have the most unique impacts.

For land managers / foresters with interest in conservation, inheriting land affected by past decisions is a reality. One of the legacies is where the land was placed. To best manage it in the future, an initial assessment of the ecological status of the land is important to understand it's likely trajectory, and what it will need in the future for tailed services/conservation/protection.

Edge effects are known for degrading primary forest, but they can also be used to assist reforestation.

Traditional edge effects act by mortality of large trees that makes room for pioneer species to enter, but existing interior pioneer species also have dispersal radii that overlap with the adjacent exterior, potentially facilitating regeneration.

Encroachment vs. facilitation relies on a balance between mortality (per area) and dispersal + establishment + stem growth.

So pioneer species can be helpful for reforestation though they are typically seen as bad for old forest conservation.

Edge effects, or changes in composition along management gradients, are likely a key variable for the sustainability of neo-tropical landscapes that are heterogenous in management (Wright et al 2009),

with underlying arguments based on metapopulation theory in ecology (Levins 1969, MacArthur & Wilson 1967), but edge effects have often been studied as threats to nature reserves, rather than how edges might facilitate secondary succession.

Especially for abandoned land that regenerates via various processes (Stanturf et al 2014), the proximity to less disturbed habitat can affect the composition of the regrowing community by changing effective seed dispersal kernels and probability of establishment (Wandrag et al 2017, Guariguata & Ostertag 2001), along with animal disperser communities themselves (Gilroy et al 2014).

While dispersal is complicated (Muller-Landau & Hardesty 2005), it can broadly correlate with simpler plant traits such as height and seed mass (Tamme et al 2014, Thomson et al 2011).

Edge effects can also operate via the style of management, including logging decisions (Edwards et al 2014, Clark et al 2012) and land-use history generally (Omeja et al 2012), can also affect stand diversity maintenance,

with examples of high diversity (compared to older adjacent stands) in agroforests (Oliveira-Neto et al 2017) and native timber plantations (Pryde et al 2015),

possibly by changing early tree survival probabilities, growth rates, and/or competition,

but surprisingly less likely facilitation by nitrogen-fixing trees (Lai et al 2018).

It may also be that due to both high recruitment and mortality (turnover) of pioneering species' stems, transitions from early- to late-successional communities may be more complex than gradual (van Greugel et al 2007).

The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density (César et al 2021), and could therefore require a switch in focal timber species (Karsten et al 2014).

To understand the ecological principles behind secondary forest regeneration (Hulshoff & Powers 2019), here we study the regeneration of an abandoned timber plantation as a function of distance from surrounding native forest.

-   how can we expect it to regrow?
-   how far/much [effort] do we need to re-seed?

We hypothesized that planted canopy species would facilitate regeneration by reducing competitive suppression of more shade-tolerant taxa by pioneering species, having mixed effects on stand diversity.

We also considered how the legacy plantation canopy structure mediated forest stand effects via changes in abiotic conditions like light availability and microtopography (Powers & Peréz-Aviles 2012).

>>>>

---
title: "Introduction"
output: html_notebook
bibliography: ../../library.bib
link-citations: TRUE
---

<!-- # Outline -->

<!-- 1.  importance of reforestation/regeneration/regrowth -->
<!-- 2.  patterns of reforestation -->
<!--     1.  general -->
<!--     2.  tropical -->
<!-- 3.  things that affect reforestation -->
<!--     1.  edge effects -->
<!--     2.  plantation structure? -->
<!-- 4.  [for NGO--start biomass census] -->
<!--     1.  Does leaving the plantation structure help reforestation? -->
<!--         1.  shade -->
<!--         2.  edge effects (i.e. buffer) -->
<!--     2.  How does the forest regenerate by distance to primary forest? -->
<!--         1.  how much [effort] do we need to re-seed? / how much can we leave to nature? -->

<!-- # Draft -->

Forest and landscape restoration is a key international conservation and climate change adaptation strategy *[@depinto20]*, and while tropical forests specifically store most land biomass *[@pan13]*, most are now secondary *[@fao20]* and regenerating degraded functions *[@hubau20]* amid insufficient policy support *[@chazdon18]*.
Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
Neotropical secondary forests can grow quickly enough to accumulate biomass >10x faster than old-growth forests *[@finegan96; @poorter16]*, and substantially offset carbon emissions, with estimates ranging from ~10% from Amazonian deforestation *[@smith20]* to a decade of fossil fuel emissions from all of Latin American and the Caribbean *[@chazdon16]*.
They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80% of old-growth forest levels in just 20 years, especially in wetter regions *[@rozendaal19]*.
The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services *[@guariguata01; @stanturf14]* like biomass and carbon storage *[@liu18]*.
However, secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*, potentially taking over a century to recover *[@poorter21]*.
Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa *[@carlo16]*, which can be keystone to locally-adapted food webs, including birds *[@maas16]*.
As a result, understanding secondary forest regeneration remains important...

Secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, as well as land-use history, especially in heterogeneously-managed landscapes *[@perfecto09; @melo13]*.
About 70% of all forests have been estimated to be within a km of their edge *[@haddad15]*, resulting in tropical forests having been estimated to lose ~22% aboveground carbon along heavily-managed edges >100m, in part due to shortening trees and lighter leaf mass *[@ordway20]*.
Edge effects also have the potential to shape community composition, based on the local dispersal kernel patterns of existing taxa *[@wandrag17]*, which while complicated *[@muller-landau09]*, can also correlate with simpler plant traits like height and seed mass *[@thomson11; @tamme14]*.
The traits of existing taxa also depend on land-use history and previous management *[@omeja12]*, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
Legacy effects from management like intensive timber planting could in theory delay or result in non-linear changes between successional stages *[@albrich20; @gough21]*, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry *[@lefland18]*.
Overall, while edge effects near degraded habitat edges are often negative, meta-population *[@levins79]* and -community *[@leibold04; @warren15]* theories highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity.
Near primary forest, dispersal may facilitate secondary forest regeneration, and help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests *[@oliveira-neto17]*, forests affected by logging *[@clark12; @edwards14]*, and timber plantations *[@pryde15]*.
<!-- role of N-fixing trees... *[@lai18]* -->

This study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, highlighting edge effects from an adjacent primary forest.
We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade that would suppress less shade-tolerant pioneer tree taxa and allow for more tree diversity closer to the primary forest edge.
    <!-- - supported by inference?! -->
Accordingly, we predicted that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.

<!-- # attic -->

<!-- Forest regeneration has the potential to store *## C* and preserve *## diversity* globally in the coming decades (UN, Bonn challenge). -->

<!-- At local scales, regeneration is strongly affected by landscape [ecological] heterogeneity as well as management history. -->

<!-- However, patterns of forest regeneration remain unclear, especially at more local scales (< ~50 ha). -->

<!-- Syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (~50-80%, compared to old-growth forests) rather quickly (decades, < 20 years), mainly with respect to tree height, above-ground biomass (Poorter et al 2016), and species richness, especially in wetter regions *[@rozendaal19; @onyekwelu16]*. -->

<!-- However, while restoring species richness may be enough to maintain some ecosystem functions like carbon storage (Liu et al 2018), it does not guarantee the restoration of species composition. -->

<!-- Instead, changes to species composition likely last up to centuries,  -->

<!-- [High] Regional biodiversity [can] affects/increases the uncertainty about the local occurrences and outcomes of specific species interactions, -->

<!-- such as after dispersal and/or niche overlap [CITE-niche, neutral theory]. -->

<!-- [At the same time,] Diverse local management activity modifies demographic states/profiles of various taxa's population dynamics. -->

<!-- For example, active timber plantations strongly promote the population growth of focal taxa at the expense of non-focal taxa. -->

<!-- After abandonment, however, the ecosystem likely reflects how continuous regional ecological processes look different given past management. -->

<!-- such as seed dispersal from nearby primary forest edges. -->

<!-- [These] Interactions between background ecology and human management ultimately have both local and general implications for the potential conservation/stability of ecosystem services/functions -->

<!-- like biomass/carbon storage and biodiversity maintenance, -->

<!-- as well as manager livelihood. -->

<!-- Slower changes in composition can also be influenced by seed banks, herbivore communities, and dispersal from the surrounding landscape, all of which may affect diverging restoration trajectories among tropical forests *(@nordon15, Christensen 2014, Munroe et al 2013, Chazdon et al 2007)*. -->

<!-- but lagging changes in composition may contribute to reduced effects of diversity on other forest functions (Griess et al 2012). -->

<!-- Restoring species composition more quickly is likely a place where active forest managers can have the most unique impacts. -->

<!-- For land managers / foresters with interest in conservation, inheriting land affected by past decisions is a reality. One of the legacies is where the land was placed. To best manage it in the future, an initial assessment of the ecological status of the land is important to understand it's likely trajectory, and what it will need in the future for tailed services/conservation/protection. -->

<!-- Edge effects are known for degrading primary forest, but they can also be used to assist reforestation. -->

<!-- Traditional edge effects act by mortality of large trees that makes room for pioneer species to enter, but existing interior pioneer species also have dispersal radii that overlap with the adjacent exterior, potentially facilitating regeneration. -->

<!-- Encroachment vs. facilitation relies on a balance between mortality (per area) and dispersal + establishment + stem growth. -->

<!-- So pioneer species can be helpful for reforestation though they are typically seen as bad for old forest conservation. -->

<!-- Especially for abandoned land that regenerates via various processes (Stanturf et al 2014), the proximity to less disturbed habitat can affect the composition of the regrowing community by changing effective seed dispersal kernels and probability of establishment (Wandrag et al 2017, Guariguata & Ostertag 2001), along with animal disperser communities themselves (Gilroy et al 2014). -->

<!-- Edge effects can also operate via the style of management, including logging decisions (Edwards et al 2014, Clark et al 2012) and land-use history generally (Omeja et al 2012), can also affect stand diversity maintenance, -->

<!-- possibly by changing early tree survival probabilities, growth rates, and/or competition, -->

<!-- but surprisingly less likely facilitation by nitrogen-fixing trees (Lai et al 2018). -->

<!-- It may also be that due to both high recruitment and mortality (turnover) of pioneering species' stems, transitions from early- to late-successional communities may be more complex than gradual (van Greugel et al 2007). -->

<!-- To understand the ecological principles behind secondary forest regeneration (Hulshoff & Powers 2019), here we study the regeneration of an abandoned timber plantation as a function of distance from surrounding native forest. -->

<!-- -   how can we expect it to regrow? -->
<!-- -   how far/much [effort] do we need to re-seed? -->

<!-- We hypothesized that planted canopy species would facilitate regeneration by reducing competitive suppression of more shade-tolerant taxa by pioneering species, having mixed effects on stand diversity. -->

<!-- We also considered how the legacy plantation canopy structure mediated forest stand effects via changes in abiotic conditions like light availability and microtopography (Powers & Peréz-Aviles 2012). -->

### 2022-05-15 article/merge.Rmd
revise intro,bibs

---
title: ""
output: 
  html_notebook
  bookdown::html_document2
    number_sections: false
bibliography: ../../library.bib
---

@@

```{r child='abstract.Rmd'}
```

# Intro

@@

# Meth

@@

# Res

```{r child="results.Rmd"}
```

# Disc

@@

# Refs

>>>>

---
title: ""
author: 
  name: Nicholas Medina
  affilitation: Brandeis University
address:
  code: Brandeis University
  address: 415 South St, Waltham, MA 02453
output: 
  # html_notebook
  bookdown::html_document2:
    number_sections: FALSE
bibliography: ../../library.bib
link-citations: TRUE
---

@@

```{r child='abs.Rmd'}
```

# Introduction

@@

# Methods

@@

# Results

<!-- # ```{r child="results.Rmd"} -->
<!-- # ``` -->

# Discussion

@@

# References

### 2022-05-15 article/methods.Rmd
revise intro,bibs

---
title: "Methods"
output: 
  # html_notebook
  bookdown::html_document2:
    number_sections: false
bibliography: ../../library.bib
---

```{r echo=F, results='hide', warning=F}
knitr::opts_chunk$set(echo = F, warning=F)
library(here)
```

@@

This study was done in a neotropical lowland rainforest on the Osa Peninsula at the Greg Gund Conservation Center **(GPS)** near the Piro Biological Station run by Osa Conservation.

(See *@taylor15* for a broader ecosystem description of the region.)

The study site was a regrowing 20 ha timber plantation of *Bombacompsis quinata* abandoned in ~1990 after the dry-adapted species from the Guanacaste region *(Pérez Cordero & Kanninen 2002)* grew poorly in very wet conditions.

This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides *(Fig \@ref(fig:map))* and a wide service road on the third NE border, with primary forest beyond it.

This census was done in 2013 during the rainy season months between June and August.

@@

Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E *(Fig \@ref(fig:map))*, using GIS software *(ArcGIS, QGIS)*.

Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.

The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *[@onyekwelu16]*.

```{r map, fig.cap="Map of conservation area on Osa Peninsula, Costa Rica."}
knitr::include_graphics(here("figs/map.png"))
# magick::image_read(path = "../figs/map.png") %>% plot()
```
<!-- ![**Fig 1**](../figs/map.png) -->

@@

The diameter of all stems >10 cm wide were recorded in each census plot.

Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.

Tree height was recorded by measuring distances to both the crown and the stem at chest height *(~2.7 m)* using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and triangulating the missing side length.

Taxa were identified with localized knowledge and field guides matching common names to Latin binomials, 
and trait information was gathered from the literature.

@@

Aboveground biomass (AGB) was estimated per stem using the allometric equation by ... *[@chave15]*.

The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by ... *[@chave05] (Ngomanda et al 2014, Alvarez et al 2012)*.

The equation version with height was used since height was accessibly measurable and has shown improved local accuracy *(Khan et al 2020, Domke et al 2012)*.

Wood density (or specific gravity) values when available also improve biomass estimates even more than height *(van Breugel et al 2011)*.

The equation used was:

*AGB = 0.0673 x (ρ D^2 H)^0.976*

with *D* in cm, *H* in m, and *p* in g cm^-3.

Measured DBHs and heights were used and wood densities (*ρ*) extracted from available literature.

Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be 0.58 following World Agroforestry Database quidelines *(worldagroforestry.org)*.

@@

Original values from 2013 were later updated to modern values using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by ... *[@réjou-méchain17]*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.

@@

Non-linear regressions were run using the function *poly()* in the *stats* `r packageVersion("stats")` R package *(R Core Team 2022)*, 
which was only recorded for stem density and richness variables.

All trees censused were included in each analysis including unidentified taxa that represented a single group in taxa-specific analyses.

Community analysis and ordination was done as a PERMANOVA with the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.

Data and code are available at *nmedina17.github.com/osa*.

>>>>

---
title: "Methods"
output: 
  # html_notebook
  bookdown::html_document2:
    number_sections: FALSE
bibliography: ../../library.bib
link-citations: TRUE
---

```{r echo=F, results='hide', warning=F, message=F}
knitr::opts_chunk$set(echo = F, warning = F, message = F); here::i_am("article/methods.Rmd")
```

@@

This study was done in a neotropical lowland rainforest on the Osa Peninsula at the Greg Gund Conservation Center (approx. `r RgoogleMaps::getGeoCode("cabo matapalo")`) near the Piro Biological Station run by Osa Conservation.
(See *@taylor15* for a broader ecosystem description of the region.)
The study site was a regrowing 20 ha timber plantation of *Bombacompsis quinata* abandoned in ~1990 after the dry-adapted species from the Guanacaste region *[@perezcordero02; @hulshof20]* grew poorly in very wet conditions.
This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \@ref(fig:map)) and a wide service road on the third NE border, with primary forest beyond it.
This census was done in 2013 during the rainy season months between June and August.

@@

Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \@ref(fig:map)), using available GIS software (*ArcGIS 10 from 2010*, [esri.com](esri.com), and *QGIS 2 from 2013*, [qgis.org](qgis.org)).
Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *[@onyekwelu16]*.

```{r map, fig.cap="Map of conservation area on Osa Peninsula, Costa Rica."}
knitr::include_graphics(here::here("figs/map.png"))
# magick::image_read(path = "../figs/map.png") %>% plot()
```
<!-- ![**Fig 1**](../figs/map.png) -->

@@

The diameter of all stems >10 cm wide were recorded in each census plot.
Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem. 
Tree height was recorded by measuring distances to both the crown and the stem at chest height *(~2.7 m)* using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and triangulating the missing side length.
Taxa were identified with localized knowledge and field guides matching common names to Latin binomials, 
and trait information was gathered from the literature.

@@

Aboveground biomass (AGB) was estimated per stem using the allometric equation by *@chave15*.
The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by *@chave05 [@alvarez12; @ngomanda14]*.
The equation version with height was used since height was accessibly measurable and has shown improved local accuracy *[@domke12; @khan20]*.
Wood density (or specific gravity) values when available also improve biomass estimates even more than height *[@vanbreugel11]*.
The equation used was: $$ {AGB} = 0.0673 (ρ D^2 H)^{0.976} $$ with diameter *D* in cm, height *H* in m, and wood density or specific gravity *p* in *g cm^-3^*.
DBHs and heights were measured in the field and wood densities were extracted from available literature. 
Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be *0.58* following World Agroforestry Database guidelines [(worldagroforestry.org)](worldagroforestry.org).

@@

Original values found and recorded were later updated to modern values using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by *@rejou-mechain17*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.

@@

Non-linear regressions were run using the function *poly()* in the *stats* `r packageVersion("stats")` R package *(R Core Team 2022)*, 
which was only recorded for stem density and richness variables. 
All trees censused were included in each analysis including unidentified taxa that represented a single group in taxa-specific analyses.
Community analysis and ordination was done as a PERMANOVA with the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
Data and code are available at [nmedina17.github.com/osa](nmedina17.github.com/osa).
